An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-05-03 paper/main.tex
More structure

\usepackage{amsthm}
\usepackage{amsmath,mathtools,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}

@@

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

@@

Parikh maps and their image appears naturally as part of many operations in
model checking and the solving of string constraints in automata-based string
solvers such as \Ostrich{}, notably in representing constraints on string
lengths. While it is possible to compute the Parikh image of any context-free
language using the method described in~\cite{generate-parikh-image}, this method
produces existentially quantified clauses and take at most linear time to
produce~\cite{muscholl-linear}. However, the resulting existentially quantified
clauses are costly to eliminate, in practice making many real-world problems
intractable. Furthermore, applications to string analysis in theorem provers
require symbolic labels to handle Unicode alphabets \Fudge{citation needed}.
Conjunctions of string constraints finally, and most crucially, lead to the
computation of Parikh images of products of regular languages. Applying the
approach in \cite{generate-parikh-image} would require the computation of the
product before the Parikh image can be computed. In several instances we have
observed, the materialisation of the product of the automata exhausts the memory
of any reasonable machine due to the exponential blow-up of the product.

Addressing these concerns, our specialised calculus for Parikh images of
products of regular languages allows us to handle symbolic transition labels
naturally, while also allowing us to interleave the computation of arbitrarily
deep products of automata with the Parikh image of their product. This allows us
to let both calculations inform each other, eliminating unnecessary work, and
pruning the size of the partial products used in the computation. Moreover, the
method can be used iteratively to, as it were, chew off bite-sized chunks of the
product, thereby avoiding the problem of memory-outs.

@@

Parikh's theorem states that ignoring order, any context-free grammar has a
letter-equivalent regular language (c.f.~\cite{construction} for a construction of
such automata from context-free grammars). However, there are languages that are
not context-free that also have semilinear images under~$\ParikhMap$ (e.g.
$\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) = \ParikhMap((abc)^*) = \#a =
\#b = \#c \land \#a \geq 0$).

@@

\contents{
  \item The Anca Muscholl paper?
  \item Describe how a "simpler" function can accelerate solving
}

@@

\section{Extending and generalising the calculus}

@@

% To generalise the calculus to calculations on products of automata, we change the main predicate take arbitrarily many automata:
% \begin{definition}
%   $\ProductPredicateInstance$ is true exactly when:
%   \begin{itemize}
%   \item $h:\: \Alphabet^* \rightarrow \SomeMonoid$, a morphism to a commutative product monoid $\SomeMonoid = \prod_{i} \SomeMonoid_i$.
%   \item $\MonoidElement$ is a vector of elements of $\SomeMonoid$.
%   \item $\AutomatonVector$ is a vector of automata.
%   \item $\Automaton = \prod_{i} \AutomatonVector_i$
%   \item each $\TransitionVec_i$ is a vector of terms such that each term correspond to a
%     transition in $\AutomatonVector_i$, and $\TransitionVectors$ is a vector of those vectors. We use the terminology $\TransitionTerm(t)$ to refer to the term of the transition vector for a given transition $t$.
%   \item $y = \sum_{t \in \Transitions} h(t)$
%   \item $\SourceTransitions(\Automaton, t)$ is the set of transitions from the
%   term automata used to produce the transition $t$ in the product automaton
%   $\Automaton$.
%   \end{itemize}
%   \end{definition}

%   First we extend $\Expand$ to generate flow equations and instances of $\Connected$ for each automaton:
%   \begin{mathpar}
%     \inferrule*[left=Expand]
%       {\AndComp{\Automaton_i \in \AutomatonVector}{\FlowEq(\Automaton_i) \land \Connected(\Automaton_i)}\ldots}
%       {\ProductPredicateInstance \land \SomeClause}
%   \end{mathpar}

%   %  x(t) = sum(e : termProductEdges(t, default=0))
% Then we introduce the rule $\Materialise$, used to compute a partial product between two terms $\Automaton_i, \Automaton_j$:
% \begin{mathpar}
%   \inferrule*[left=Materialise, 
%   right=\textnormal{where $\Automaton_{k} = \Automaton_i \times \Automaton_j, \AutomatonVector' = (\AutomatonVector \cup \{  \Automaton_k \}) \setminus {\Automaton_i, \Automaton_j}$}]
%     {\AndComp{t \in \delta(\Automaton_{k})}{\TransitionTerm(t) = \sum_{t' \in \SourceTransitions(t, \Automaton_k)} t'}\\
%     \FlowEq(\Automaton_{k}) \land \Connected(\Automaton_{k}) \land
%     \\ \MonoidMap(\AutomatonVector', \TransitionVectors', \MonoidElement) \land \SomeClause} {\MonoidMap(\AutomatonVector, \TransitionVectors, \MonoidElement) \land \SomeClause}
% \end{mathpar}

% Note that we implicitly existentially quantify the fresh transition terms $\TransitionVec_{k+1}$ for the product.

% $\Materialise$ comes with its companion rule $\RemoveTermConnected$ that simply
% states we are allowed to remove any instances of $\Connected$ from terms that
% have been used in a product, and whose connectedness we therefore no longer care about:
% \begin{mathpar}
%   \inferrule*[left=$\RemoveTermConnected$, right=\textnormal{If $\Automaton_a$ not in $\Automaton_1\ldots\Automaton_k$}]
%     {\ProductPredicateInstance \land \SomeClause}
%     {\Connected(\Automaton_a) \land \ProductPredicateInstance \land \SomeClause}
% \end{mathpar}

% Finally, we add a side condition to $\Subsume$ to only allow subsumption for
% instances of $\MonoidMap$ with just one automaton, ensuring that we do not bail
% out before having computed the entire product.

\contents{
    \item additional rules needed for products
    }

@@

@@

\contents{
  \item Listing: a nice input example
  \item Do we need to say any more about the nuxmv backend?
  \item Optimisations and tricks
}

@@

\subsection{\Catra{} is Versatile: Encoding Epistemic Logic Problems}

@@

\contents{
\item length constraints
\item Parikh automata, automata with registers
\item Where do the instances come from!? Why?
\item Result table
\item Definition of winning
}

>>>>

\usepackage{amsthm}
\usepackage{amsmath,mathtools,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}

@@

%% Keywords
%% comma separated list
\keywords{parikh images, string solvers, model checking}  %% \keywords are mandatory in final camera-ready submission

@@

The Parikh map is a characterisation of a context-free language as a vector of
its character counts, famously a semilinear set~\cite{parikh-theorem}. It
appears naturally as part of many operations in model checking and the solving
of string constraints in automata-based string solvers such as \Ostrich{},
notably in representing constraints on string lengths.

It is possible to compute the Parikh image of any context-free language using
the method described in~\cite{generate-parikh-image}. Later improvements have
produced a construction taking at most linear time to
produce~\cite{muscholl-linear}. However, the resulting existentially quantified
clauses are costly to eliminate, in practice making many real-world problems
intractable.

In addition to costly elimination of the existentially quantified variables,
another problem arises for the common case of solving length constraints in
string solvers. Conjunctions of string constraints lead to the computation of
Parikh images of products of regular languages. Applying the approach
in~\cite{generate-parikh-image} would require the computation of the product
before its Parikh image. In several instances we have observed while solving
real-world string constraints, the computation the product of automata exhausts
the memory of any machine due to the exponential blow-up in size of the product,
quickly becoming intractable as the number of terms in the product increases.

Addressing these concerns, our specialised calculus for Parikh images of
products of regular languages allows us to interleave the computation of
arbitrarily deep products of automata with the product's Parikh image. This
allows us to let both calculations inform each other, eliminating unnecessary
work, and pruning the size of the partial products considered in the
computation. Moreover, the method can be used iteratively to tackle smaller
chunks of the product incrementally, thereby avoiding running out of memory.

@@

Parikh's theorem states that ignoring order, any context-free grammar has a
letter-equivalent regular language (c.f.~\cite{construction} for a construction of
such automata from context-free grammars and~\cite{bounds} for bounds on its size). However, there are languages that are
not context-free that also have semilinear images under~$\ParikhMap$ (e.g.
$\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) = \ParikhMap((abc)^*) = \#a =
\#b = \#c \land \#a \geq 0$).

@@

\contents{
  \item The Anca Muscholl paper?
  \item Describe how a "simpler" function can accelerate solving
  \item Describe the relationship of this generalisation to Parikh matrices!
}

@@

@@

To generalise the calculus to calculations on products of automata, we change the main predicate to take arbitrarily many automata:
\begin{definition}
  $\ImagePredicate{\Automaton_1\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}$
  is true exactly when the single-automaton version of the predicate would hold
  for the automaton.

  \contents{
    \item Is the enumeration correct?
    \item Do we need a new name for the predicate?
    \item Should the automata still be subscripted?
  }
\end{definition}

\begin{definition}
When computing a product of two automata, $\Automaton_1 \times \Automaton_2$, we
need to be able to reason about the sources of a transition. For that, we
introduce the relation $\ComesFrom(\Transition, \Transition',
\Automaton_1\times\Automaton_2)$. The predicate holds when $\Transition \in
\Transitions_{\Automaton_2 \times \Automaton_2}$ is created from $\Transition', \Transition''
\in \Transitions_{\Automaton_1} \cup \Transitions_{\Automaton_2}$.\Fudge{Do I need to be more specific?}
\end{definition}

  First we extend $\Expand$ to generate flow equations and instances of $\Connected$ for each automaton:
  \begin{mathpar}
    \inferrule*[left=ExpandM]
      {\Set{\Automaton_i}{\FlowEq(\Automaton_i), \Connected(\Automaton_i)}, \ImagePredicate{\Automaton_1\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause}
      {\ImagePredicate{\Automaton_1\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause}
  \end{mathpar}

\ExpandM{} must be applied before any other rule, like \Expand{}, but unlike \Expand{}, \ExpandM{} does not remove the $\Image$~predicate since it is needed to keep track of the product.

Then we introduce the rule $\Materialise$, used to compute a partial product between two terms:
\begin{mathpar}
  \inferrule*[left=Materialise,right=\textnormal{where $\Automaton' = \Automaton_1\times\Automaton_2$}]
    {\Set{\Filter(\Transition) = \sum \Filter(\Transition') \HoldsThat \Transition' \in \Transitions_{\Automaton'}, \ComesFrom(\Transition', \Transition, \Automaton') \SuchThat \Transition \in \Transitions_{\Automaton_1} \cup \Transitions_{\Automaton_2}},\\\ImagePredicate{\Automaton'\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause}
    {\ImagePredicate{\Automaton_1\times\Automaton_2\times\ldots\times\Automaton_k}{\Map}{\Filter}{\MonoidElement}, \SomeInequalities, \SomeClause}
\end{mathpar}

Note that we implicitly map $\Filter$ to fresh terms for each transition in the
product just like before.

Finally, for instances of precisely one automaton, neither rule applies and we
perform the calculus as before.

@@

\subsection{The extended calculus is also correct}
\subsubsection{The calculus terminates}
\subsubsection{The calculus allows no values outside the Parikh image}
\subsubsection{The calculus excludes no values in the Parikh image}

@@

\begin{lstlisting}[caption={An example input file for \Catra{}, illustrating every major syntax element. From beginning to end: synchronised (product) automata using the keyword \texttt{synchronised} (automata A and B), range labels, and their single-character shorthand syntax, register increments, and constraints on the final values of their counters.}, label=lst:input-example]
  /* Declare counters. Initially only integers, but other data-types
 could be considered as well. All counters start at zero. */
counter int x, y, z, length;

// List automata. Automata transitions are labelled with intervals,
// and can specify counter increments/decrements. A counter can only
// be used in at most one automaton.

synchronised {
    automaton A {
    init s0;

    s0 -> s1 [0, 10] { x += 1};

    accepting s1, s2;
    };

    automaton B {
    init t1;

    t1 -> t2 [11, 22] { y += 1};

    accepting t1, t2;
    };
};

automaton C {
    init s0;

    s0 -> s1 [15] { z += 1 };
    s1 -> s1 [15] { z += 1};

    accepting s1;
};

// Specify a constraint on the final values of counters; check whether
// there is a word accepted by all automata for which the final counter
// values satisfy this constraints.
constraint z = 2x + 2y && !(x = 0);
\end{lstlisting}

\contents{
  \item Do we need to say any more about the nuxmv backend?
  \item Optimisations and tricks
}

@@

\subsection{\Catra{} is Versatile: Encoding Epistemic Logic Problems As Parikh Automata}

@@

\contents{
\item length constraints
\item Parikh automata, automata with registers
\item Where do the instances come from!? Why?
\item Result table
\item Definition of winning
\item We compare to an implementation using the same automated theorem prover, the same product implementation
}
